Neutral models of \textit{de novo} gene emergence show that gene evolution has a preferred trajectory

### 2022-11-09 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex

We reframed our question to make it more relatable to commonly observed findings from genomics experiments. For example, a hypothetical experiment involves DNA sequencing of a genomic locus in different related species. This experiment shows that transcription features are present in only one species, whereas the ORF is present in all species. Two evolutionary events could describe this observation. First possibility is that a \textit{de novo} gene existed in an ancestor and transcription was lost in multiple species. For any one species, the probability of this evolutionary trajectory is: $P_\textit{ORF-stay}\times P_\textit{RNA} \times P_\textit{RNA-loss}$. (Note that $P_\textit{RNA-loss}$ is a conditional probability, that is dependent on the presence of transcription).

@@

Alternatively, an ORF could have existed in all species and transcription was subsequently gained in one species. The probability that this event happens in any one species is: $P_\textit{ORF-stay}\times P_\textit{RNA-gain}$ (see \autoref{orffirsteq1}).

The ratio of the probabilities of the two evolutionary trajectories would simply be: $P_\textit{RNA} \times P_\textit{RNA-loss}/P_\textit{RNA-gain}$. We found the value of this ratio to be 0.914, suggesting that transcription loss in a \textit{de novo} gene would be more probable than transcription gain for an untranscribed ORF. However, this difference is small and hence the real evolutionary mechanism behind the observation from our hypothetical sequencing experiment, would be difficult to infer.

In sum, transcription loss cannot be reliably inferred as an evidence for negative selection of protein expression.

@@

Like all theoretical models, our model is not without limitations. First, we assume that nucleotides are uniformly distributed according to GC content. This is not true as nucleotide composition can vary significantly across the genome. Mutation rates are also not uniform across the genome and can vary significantly \citep{mutbiasArabidopsis}. However, our model can be adapted to investigate specific genomic loci, where information on the local GC content and mutation rate can more accurately determine the \textit{de novo} gene evolution in these loci. Second, our model is based on parameters such as mutation rate and mutation bias, which we obtained from published data from humans and \textit{D.melanogaster} \citep{humanmutrate,drosophilamutrate}. The accuracy of the model will depend on the accuracy of parameters. More importantly, different species may have a different spectrum of mutations \citep{joshmutbias}, especially so for species that are highly distant from each other. Thus our inferences based on our chosen parameters will not universally apply to all species. Third, we assume that all mutations are independent. This is not the case for tandem mutations, where multinucleotide mutations occur can more frequently than their expected frequency under the assumption of mutational independence \citep{MNM}. Finally, we ignore insertions, deletions and transpositions as mechanisms of mutation. Insertions and deletions (indels) are reported to be less frequent than substitutions \citep{drosophilamutrate}, and both indels transpositions cause larger sequence alterations than point mutations. However, our modeling methodology allows incorporation of these mechanisms as a source of mutations.

>>>>

We made our analysis of gene loss more stringent by comparing the probabilities of two evolutionary scenarios, both of which begin with an intact \textit{de novo} gene. In the first scenario, the ORF remains intact whereas the transcription is lost. The probability of this scenario is described by: 
$P_\textit{ORF-stay} \times P_\textit{RNA-loss}/P_\textit{ORF}$. The probability of the second scenario, where transcription stays intact while ORF is lost, is: $P_\textit{ORF-loss} \times P_\textit{RNA-stay}/P_\textit{RNA}$. We find that the first scenario is more plausible for ORFs that have less than 34 codons, which is in agreement with our previous analysis (\autoref{lossprob}). Also, the probability values of scenario one and two, obtained from this stringent analysis, were not significantly different from the transcription ($P_\textit{RNA-loss}$) and ORF loss ($P_\textit{ORF-loss}$) probabilities, respectively.

Thus widespread transcription loss can suggest a negative selection of

@@

@@

Like all theoretical models, our model is not without limitations. First, we assume that nucleotides are uniformly distributed according to GC content. This is not true as nucleotide composition can vary significantly across the genome. Mutation rates are also not uniform across the genome and can vary significantly \citep{mutbiasArabidopsis}. However, our model can be adapted to investigate specific genomic loci, where information on the local GC content and mutation rate can more accurately determine the \textit{de novo} gene evolution in these loci. Second, our model is based on parameters such as mutation rate and mutation bias, which we obtained from published data from humans and \textit{D.melanogaster} \citep{humanmutrate,drosophilamutrate}. The accuracy of the model will depend on the accuracy of parameters. Thus our inferences based on our chosen parameters will not universally apply to all species. However, mutation rate itself only determines the values of different probabilities. Our results, which are based on probability ratios, will not significantly change with a different value of mutation rate. More importantly, different species may have a different spectrum of mutations \citep{joshmutbias}, especially so for species that are highly distant from each other. Third, we assume that all mutations are independent. This is not the case for tandem mutations, where multinucleotide mutations occur can more frequently than their expected frequency under the assumption of mutational independence \citep{MNM}. Finally, we ignore insertions, deletions and transpositions as mechanisms of mutation. Insertions and deletions (indels) are reported to be less frequent than substitutions \citep{drosophilamutrate}, and both indels transpositions cause larger sequence alterations than point mutations. However, our modeling methodology allows incorporation of these mechanisms as a source of mutations.
